---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

## Loading and preprocessing the data

Load in the data.
```{r}
unzip("activity.zip")
act <- read.csv("activity.csv")
```

Load required libraries.
```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
```

Explore the data to determine preprocessing needed.
```{r}
# str(act)        # date needs to be formatted
# summary(act)    # NAs are present in data frame
act$date <- as.Date(act$date, "%Y-%m-%d")
act_noNA <- act[!is.na(act$steps),]
```


## What is mean total number of steps taken per day?
```{r}
totalStepsPerDay <- tapply(act_noNA$steps, act_noNA$date, FUN=sum)
```

The total no. of steps taken per day is:
```{r}
totalStepsPerDay
```

Histogram of the total no. of steps taken each day.
```{r}
# range(totalStepsPerDay)/30               # to determine suitable bin width
qplot(totalStepsPerDay, binwidth=500)
```

The mean total no. of steps per day is: `r mean(totalStepsPerDay)` (`mean(totalStepsPerDay)`)  
The median total no. of steps per day is: `r median(totalStepsPerDay)` (`median(totalStepsPerDay)`)


## What is the average daily activity pattern?

```{r}
act_grpByIntervals <- act_noNA %>%
                      group_by(interval) %>%
                      summarise(meanSteps=mean(steps))

g <- ggplot(act_grpByIntervals, aes(x=interval, y=meanSteps))
g <- g + geom_line()
g
```

The 5-minute interval that contains the max. no. of steps on average: `r act_grpByIntervals[which.max(act_grpByIntervals$meanSteps),1]` (`act_grpByIntervals[which.max(act_grpByIntervals$meanSteps),1]`)


## Imputing missing values

The no. of rows with missing data:
```{r}
which(is.na(act)) %>% unique() %>% length()
```

The means of the respective 5-minute intervals is used to replace the missing data.
```{r}
rowIdsForNAs <- which(is.na(act$steps))
intervalsAtNAs <- act$interval[rowIdsForNAs]

act_simulatedNAs <- act                 # make replacement on new df

act_simulatedNAs$steps[rowIdsForNAs] <- act_grpByIntervals$meanSteps[
                                        match(intervalsAtNAs, act_grpByIntervals$interval)]

totalStepsPerDaySimulated <- tapply(act_simulatedNAs$steps, act_simulatedNAs$date, FUN=sum)
```

Histogram of the total no. of steps taken each day.
```{r}
# range(totalStepsPerDaySimulated)/30               # to determine suitable bin width
qplot(totalStepsPerDaySimulated, binwidth=500)
```

The mean total no. of steps per day is: `r mean(totalStepsPerDaySimulated)` (`mean(totalStepsPerDaySimulated)`)  
The median total no. of steps per day is: `r median(totalStepsPerDaySimulated)` (`median(totalStepsPerDaySimulated)`)

After imputing missing values:  
1. The mean total no. of steps per day remains the same;  
2. The median total no. of steps per day has increased towards to mean.


## Are there differences in activity patterns between weekdays and weekends?

A new column with factors "Weekend" and "Weekday" is introduced.
```{r}
act_ByDayOfWeek <- act_simulatedNAs     # to introduce new factor column
act_ByDayOfWeek$Day <- weekdays(act_ByDayOfWeek$date, abbreviate=TRUE)
act_ByDayOfWeek$Day <- ifelse(act_ByDayOfWeek$Day %in% c("Sat", "Sun"), "Weekend", "Weekday")
act_ByDayOfWeek$Day <- as.factor(act_ByDayOfWeek$Day)

act_ByDayOfWeekGrp <- act_ByDayOfWeek %>%
                      group_by(interval, Day) %>%
                      summarise(meanSteps=mean(steps))
```

Panel plot of average no. of steps taken by 5-minute intervals, faceted by "Weekend" and "Weekday"
```{r}
g2 <- ggplot(act_ByDayOfWeekGrp, aes(x=interval, y=meanSteps))
g2 <- g2 + facet_grid(Day ~ .) + geom_line()
g2
```

